---
title: CCHS Ontario population for microsimulations
author: Hana Fu
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    number_sections: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

```


```{r lib, message=FALSE, warning=FALSE}

library(tidyverse)
library(kableExtra)
library(ggplot2)

```

```{r documentations}

# https://www150.statcan.gc.ca/n1/pub/12-002-x/2014001/article/11901-eng.htm
# http://digrs.blogspot.com/2014/12/cchs-bootstrap-weights.html
# https://gsg.uottawa.ca/data/teaching/escc-cchs/fmgd_pumf/2015-2016/bootstrap_e.pdf
# http://www.ciqss.umontreal.ca/Docs/Formations/Ateliers/2003-11-27_UseofData.pdf
# http://gsg.uottawa.ca/data/eds/Bootvar/AppendixC.pdf (CCHS bootstrap weights don't seem to be mean bootstrap)


```

```{r load_data}

dataDir <- 'H:\\Microsimulations\\Data'

load(file.path(dataDir,'cchs2013_14_Ontario.RData'))

saveRDS(cchs2013_14_Ontario[1:200,], file.path(dataDir,'cchs2013_14_Ontario2.rds'))


```

# Sample and bootstrap weights

Histogram display the sum of each 500 bootstrap weights, with <span style='color: blue;'><b>blue</b></span> line indicates the sum of sample weights (=`r format(sum(cchs2013_14_Ontario$WTS_S), big.mark= ",", scientific = FALSE)`). This is to check whether bootstrap weights already include the sample weights, which seem to be true.

```{r some_checks}

# Some checks on variables
#sum(table(cchs2013_14_Ontario$SAMPLEID) > 1) # should be zero, that is, each sample ID does not appear more than once

# Sum of sample weights
sampleWgtSum <- sum(cchs2013_14_Ontario$WTS_S)

# Sum of bootstrap weights
btstrpVars <- grep("BSW[0-5]+", names(cchs2013_14_Ontario), value = TRUE)
btstrpSum <- colSums(cchs2013_14_Ontario[btstrpVars])
btstrpSum <- data.frame(btstrpSum)
colnames(btstrpSum) <- "value"

# Plot
ggplot(btstrpSum, aes(x=value)) +
  geom_histogram(colour="white") +
  geom_vline(xintercept = sampleWgtSum, 
                color = "blue", size=1.5)

```

<!-- # Create factor variables -->

```{r cchs_vars, echo=TRUE}

cchs2013_14_Ontario$sex <- factor(cchs2013_14_Ontario$DHH_SEX, labels=c("Male","Female"))

cchs2013_14_Ontario$agecat <- cut(cchs2013_14_Ontario$DHH_AGE, c(12,30,40,50,60,70,80,105), 
                                  include.lowest = TRUE, right = FALSE)
levels(cchs2013_14_Ontario$agecat) <- c("12-29","30-39","40-49","50-59","60-69","70-79","80+")

cchs2013_14_Ontario$educat <- factor(cchs2013_14_Ontario$EDUDR04, 
                                     labels=c("Less than secondary","Secondary school grad", 
                                     "Some post-secondary", "Post-secondary grad", "NS"),
                                     levels =c(1,2,3,4,9))

cchs2013_14_Ontario$incpercat <- with(cchs2013_14_Ontario,
                                      ifelse(INCDPER < 10 | INCDPER %in% c(96,99), INCDPER,
                                             ifelse(INCDPER %in% c(10,11), 10,
                                                    ifelse(INCDPER %in% c(12,13), 11, 
                                                           ifelse(INCDPER == 14, 12, NA)))))

cchs2013_14_Ontario$incpercat <- factor(cchs2013_14_Ontario$incpercat,
                                        labels = c("No income", "LESS THAN $5,000", "$5,000 TO $9,999", "$10,000 TO $14,999", "$15,000 TO $19,999", "$20,000 TO $29,999", "$30,000 TO $39,999", "$40,000 TO $49,999", "$50,000 TO $59,999", "$60,000 TO $79,999","$80,000 TO 99,999","$100k or more","NS"),
                                        levels =c(1:12,99))

cchs2013_14_Ontario$inchhcat <- with(cchs2013_14_Ontario,
                                      ifelse(INCDHH < 10 | INCDHH %in% c(96,99), INCDHH,
                                             ifelse(INCDHH %in% c(10,11), 10,
                                                    ifelse(INCDHH %in% c(12,13), 11, 
                                                           ifelse(INCDHH %in% c(14,15), 12, NA)))))
cchs2013_14_Ontario$inchhcat <- factor(cchs2013_14_Ontario$inchhcat,
                                        labels = c("No income", "LESS THAN $5,000", "$5,000 TO $9,999", "$10,000 TO $14,999", "$15,000 TO $19,999", "$20,000 TO $29,999", "$30,000 TO $39,999", "$40,000 TO $49,999", "$50,000 TO $59,999", "$60,000 TO $79,999","$80,000 TO 99,999","$100k or more","NS"),
                                        levels =c(1:12,99))

cchs2013_14_Ontario$ethncat <- factor(cchs2013_14_Ontario$SDCDCGT,
                                      labels = c('White','Black','Korean','Filipino','Japanese','Chinese','South Asian','South East Asian','Arab','West Asian','Latin American','Other Racial or Cultural Origin','Multiple Racial/Cultural Origins','NA','NS'),
                                      levels = c(1:13,96,99))

cchs2013_14_Ontario$maritalcat <- factor(cchs2013_14_Ontario$DHH_MS,
  labels = c('Married','Common-Law','Widowed','Separated','Divorced','Single/Never Married','DK','Refusal','NS'),
  levels = c(1:6,97,98,99))

cchs2013_14_Ontario$geourcat <- factor(cchs2013_14_Ontario$GEODUR2,
       labels=c("Population Centre", "Rural"),
       levels=c(1,2))

cchs2013_14_Ontario$alcfreqcat <- factor(cchs2013_14_Ontario$ALC_2,
                                         labels = c('less than once a month','once a month','2 to 3 times a month','once a week','2 to 3 times a week','4 to 6 times a week','every day','NA','DK','Refusal','NS'),
                                         levels = c(1:7,96:99))

cchs2013_14_Ontario$alctypecat <- factor(cchs2013_14_Ontario$ALCDTTM,
                                         labels = c("Regular drinker","Occasional drinker","Did not drink in the last 12 months","NS"),
                                         levels = c(1,2,3,9))

cchs2013_14_Ontario$alchvycat <- with(cchs2013_14_Ontario,
                                      ifelse(ALC_3 %in% 1:2, 2, # 1 is heavy drinking
                                        ifelse(ALC_3 %in% 3:6, 1, # 2 did not have heaving drinking
                                               ifelse(ALC_3 %in% 96:99, ALC_3, NA))))

cchs2013_14_Ontario$alchvycat <- factor(cchs2013_14_Ontario$alchvycat,
                                        labels = c("Heavy drinking in the past year at least once per month","Did not heavy drink in the past year","NA","DK","Refusal","NS"),
                                        levels = c(1:2,96:99))

```


# Tabulation of counts by age and sex (without survey weights)

```{r crude_tab}
# Crude tabulation
crudeTab <- cchs2013_14_Ontario %>%
  group_by(agecat, sex) %>%
  summarise(count = n()) %>%
  pivot_wider(names_from = sex, values_from = count, values_fill = 0) %>%
  mutate(Both = Male + Female)
crudeTotal <- data.frame(agecat="Total", Male = sum(crudeTab[,2]), 
                         Female = sum(crudeTab[,3]), Both = sum(crudeTab[,4]))
# Print tabulation
rbind(crudeTab, 
      crudeTotal) %>%
  kable(col.names = c("Age group","Male","Female","Both")) %>%
  kable_styling("striped") %>%
  row_spec(8, bold = TRUE)
```

# Tabulation of counts by socioeconomic variables (without survey weights)

```{r crude_ses_tab, cache=TRUE}

# Crude socioeconomic tabulation
crudeSesTab <- cchs2013_14_Ontario %>%
  group_by(agecat, sex, educat, incpercat, inchhcat, ethncat, maritalcat, geourcat) %>%
  summarise(count = n()) 

# Print tabulation
DT::datatable(crudeSesTab, colnames=c("Age group","Sex","Education","Income (personal)","Income (household)",
                      "Ethnicity","Marital Status","Urban/Rural", "Count"),
              filter = 'top',
              options = list(pageLength = 20))

```

# Tabulation of counts by alcohol drinking status (without survey weights)

```{r crude_alc_tab}

# Crude alcohol tabulation
crudeAlcTab <- cchs2013_14_Ontario %>%
  group_by(alcfreqcat, alctypecat, alchvycat) %>%
  summarise(count = n()) 

# Print tabulation
DT::datatable(crudeAlcTab, colnames=c("Frequency of drinking alcohol", "Type of drinker","Heavy drinking", "Count"),
              filter = 'top',
              options = list(pageLength = 20))

```

# Implement survey design

```{r svy_design, echo=TRUE, message=FALSE, warning=FALSE}

# See link for instructions https://stackoverflow.com/questions/45642026/survey-weights-and-boostrap-wieghts-to-get-counts-and-cis

library(survey)
cchsDesign <- svrepdesign(data = cchs2013_14_Ontario,
                        weights = ~WTS_S,
                        repweights = "BSW[0-5]+",
                        type = "bootstrap",
                        combined.weights = TRUE,
                        mse = TRUE)

```

## Obtain survey-weighted population size

### Age and sex

```{r svy_tab, echo=TRUE}

svyTabRes <- svytable(
  ~agecat + sex, # variable to pass to function
  design = cchsDesign # design object
  )

```

```{r print_svy_tab}

# Tabulation
svyTab <- svyTabRes %>%
  data.frame %>%
  pivot_wider(names_from = sex, values_from = Freq, values_fill = 0) %>%
  mutate(Both = Male + Female)

svyTotal <- data.frame(agecat="Total", Male = sum(svyTab[,2]), 
                         Female = sum(svyTab[,3]), Both = sum(svyTab[,4]))

# Print tabulation
rbind(svyTab, 
      svyTotal) %>%
  kable(col.names = c("Age group","Male","Female","Both")) %>%
  kable_styling("striped") %>%
  row_spec(8, bold = TRUE)

```

### Socioeconomic and alcohol drinking status

Here, eight variables are included to generate survey-weighted counts. The eight variables are age, sex, education, household income, ethnicity, marital status, urban/rural setting, and heavy drinking.

```{r svy_ses_tab, echo=TRUE}

svySesAlcTabRes <- svytable(
  ~agecat + sex + educat + inchhcat + ethncat + maritalcat + geourcat + alchvycat, # variable to pass to function
  design = cchsDesign # design object
  )

```

The codes above creates a 8 dimension output in the "svySesAlcTabRes" object. Many of the "slices" contains zeros, here are a couple of examples:

> svySesAlcTabRes[,,"Secondary school grad","LESS THAN $5,000","Black","Married","Population Centre","Heavy drinking in the past year at least once per month"]

```{r zero_slice_1}

svySesAlcTabRes[,,"Secondary school grad","LESS THAN $5,000","Black","Married","Population Centre","Heavy drinking in the past year at least once per month"]
       
```

> svySesAlcTabRes[,,"Less than secondary","No income","White","Single/Never Married","Rural","Did not heavy drink in the past year"]

```{r zero_slice_2}

svySesAlcTabRes[,,"Less than secondary","No income","White","Single/Never Married","Rural","Did not heavy drink in the past year"]

```

There are also "slices" that are non-zero:

> svySesAlcTabRes[,,"Post-secondary grad","$80,000 TO 99,999","White","Married","Population Centre","Heavy drinking in the past year at least once per month"]

```{r non_zero_slice}

svySesAlcTabRes[,,"Post-secondary grad","$80,000 TO 99,999","White","Married","Population Centre","Heavy drinking in the past year at least once per month"]

```


# Tabulation of counts by selected socioeconomic variables with survey weights (without using survey package)

```{r crude_ses_tab_for_sim, cache=TRUE}

# Crude socioeconomic tabulation
crudeSes3Tab <- cchs2013_14_Ontario %>%
  group_by(sex, educat, geourcat) %>%
  summarise(count = n()) 

# Print tabulation
DT::datatable(crudeSes3Tab, colnames=c("Sex","Education","Urban/Rural", "Count"),
              filter = 'top',
              options = list(pageLength = 20))

# Assign strata to the cCHS data.frame
# assignStrata() function returns a data.frame with the stratum and stratum.name columns, by providing the categorical variables and specific categories to remove for variables
variable.names <- c("sex","educat","geourcat")
remove.categories <- c("NA","DK","Refusal","NS")

assignStrata <- function(data, variable.names, remove.categories) {
  
  crudeTab <- data %>% 
    group_by_at(variable.names) %>%
    summarise(count = n())
  
  removeRow <- apply(crudeTab[variable.names], 1, function(i) {
    sum(!is.na(match(i, remove.categories))) == 1 # if the stratum contains any of the remove.categories, it returns as TRUE, else FALSE
  })
  
  crudeTab <- crudeTab[!removeRow,]
  
  for (i in 1:nrow(crudeTab)) {
    crudeTabSub <- crudeTab[i,variable.names]
    
    stratumMat <- matrix(NA, nrow=nrow(data),
           ncol=length(variable.names),
           dimnames = list(1:nrow(data), variable.names))
    
    for (var in variable.names) {
      stratumMat[,var] <- data[,var] == unlist(crudeTabSub[,var])
    }
    
    all.true <- apply(stratumMat, 1, all)
  
    data[which(all.true),'stratum'] <- i 
    data[which(all.true),'stratum.name'] <- paste(apply(crudeTabSub,2,as.character), collapse=':') 
  }
  
  return(list(data = data, crudeTab = crudeTab))
}

strataOutput <- 
 assignStrata(cchs2013_14_Ontario, variable.names, remove.categories)

cchs2013_14_Ontario_strata <- strataOutput$data
crudeTab <- strataOutput$crudeTab

cchs2013_14_Ontario_strata %>% select(sex, agecat, educat, incpercat, inchhcat, ethncat, maritalcat,
                                      geourcat, alcfreqcat, alctypecat, alchvycat, stratum, stratum.name)

# Need to provide transition probabilities for the strata
# This is in matrix format (rows for strata and columns for the transition probabilities)
# For now, just generate the transition probabilities using runif()
transProbMat <- matrix(NA, nrow=nrow(crudeTab), ncol=length(variable.names)+5,
                       dimnames = list(1:nrow(crudeTab), c(variable.names,'p.HD','p.HS','p.SH','rr.S','rp.S')))
# p.HD: probability to die when healthy
# p.HS: probability to become sick when healthy
# p.SH: probability to become healthy when sick
# rr.S: rate ratio of death when sick vs healthy
# rp.s: increase of the mortality rate

p.HD <- 0.005
p.HS <- 0.15 
p.SH <- 0.5 
rr.S <- 3 
rp.S <- 0.2 
# The transition probabilities and rr.S are updated based on sex, education, and geography; rp.S remains the same
# Male and rural area has higher probability (+0.005) than female and urban area (-0.001) for healthy to death
# Male and rural area has higher probability (+0.01) than female urban area (-0.01) for healthy to sick
# Male and rural area has higher probability (-0.05) than female urban area (+0.05) for sick to health
# Male and rural area has higher probability (+0.25) than female and urban area (-0.25) for rate ratio of death when sick vs healthy
# For education, less than secondary (+0.075), secondary school grad (+0.05), Some post-secondary (+0.025) for healthy to death
# For education, less than secondary (+0.1), secondary school grad (+0.075), Some post-secondary (+0.05) for healthy to sick
# For education, less than secondary (+0.1), secondary school grad (+0.075), Some post-secondary (+0.05) for sick to healthy
# For education, less than secondary (+0.25), secondary school grad (+0.15), Some post-secondary (+0.05) for rate ratio of death when sick vs healthy
crudeTab$p.HD_sexAdj <- with(crudeTab, ifelse( sex == "Male", 0.005, -0.001))
crudeTab$p.HD_geoAdj <- with(crudeTab, ifelse( geourcat == "Rural", 0.005, -0.001))
crudeTab$p.HS_sexAdj <- with(crudeTab, ifelse(sex == "Male", 0.01, -0.01))
crudeTab$p.HS_geoAdj <- with(crudeTab, ifelse( geourcat == "Rural", 0.01, -0.01))
crudeTab$p.SH_sexAdj <- with(crudeTab, ifelse( sex == "Male", -0.05, 0.05))
crudeTab$p.SH_geoAdj <- with(crudeTab, ifelse( geourcat == "Rural", -0.05, 0.05))
crudeTab$rr.S_sexAdj <- with(crudeTab, ifelse( sex == "Male", 0.25, -0.25))
crudeTab$rr.S_geoAdj <- with(crudeTab, ifelse( geourcat == "Rural", 0.25, -0.25))

crudeTab$p.HD_eduAdj <- with(crudeTab, ifelse( educat == "Less than secondary", 0.075,
                               ifelse(educat == "Secondary school grad", 0.05,
                                      ifelse(educat == "Some post-secondary", 0.025, 0))))
crudeTab$p.HS_eduAdj <- with(crudeTab, ifelse( educat == "Less than secondary", 0.1,
                               ifelse(educat == "Secondary school grad", 0.075,
                                      ifelse(educat == "Some post-secondary", 0.05, 0))))
crudeTab$p.SH_eduAdj <- with(crudeTab, ifelse( educat == "Less than secondary", 0.1,
                               ifelse(educat == "Secondary school grad", 0.075,
                                      ifelse(educat == "Some post-secondary", 0.05, 0))))
crudeTab$rr.S_eduAdj <- with(crudeTab, ifelse( educat == "Less than secondary", 0.25,
                               ifelse(educat == "Secondary school grad", 0.15,
                                      ifelse(educat == "Some post-secondary", 0.05, -0))))

crudeTab$p.HS <- p.HS + with(crudeTab, p.HS_sexAdj + p.HS_eduAdj + p.HS_geoAdj)
crudeTab$p.HD <- p.HD + with(crudeTab, p.HD_sexAdj + p.HD_eduAdj + p.HD_geoAdj)
crudeTab$p.SH <- p.SH + with(crudeTab, p.SH_sexAdj + p.SH_eduAdj + p.SH_geoAdj)
crudeTab$rr.S <- rr.S + with(crudeTab, rr.S_sexAdj + rr.S_eduAdj + rr.S_geoAdj)
crudeTab$rp.S <- rp.S

crudeTab[c(variable.names,'p.HS','p.HD','p.SH','rr.S','rp.S')]

transProbMat <- crudeTab[c(variable.names, 'p.HS','p.HD','p.SH','rr.S','rp.S')]

```