---
title: CCHS Ontario population for microsimulations
author: Hana Fu
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    number_sections: true
---
  
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

```


```{r lib, message=FALSE, warning=FALSE}

library(tidyverse)
library(kableExtra)
library(ggplot2)

```

```{r load_data}

dataDir <- 'D:\\Microsimulations\\Data'

load(file.path(dataDir,'cchs2013_14_Ontario.RData'))

saveRDS(cchs2013_14_Ontario[1:200,], file.path(dataDir,'cchs2013_14_Ontario2.rds'))

```


```{r wts, eval=FALSE}

idx <- grep('WTS_S|BSW', names(cchs2013_14_Ontario))
table(round(colSums(cchs2013_14_Ontario[idx])))

```

<!-- # Create factor variables -->
  
```{r cchs_vars, echo=TRUE}

cchs2013_14_Ontario$sex <- factor(cchs2013_14_Ontario$DHH_SEX, labels=c("Male","Female"))

cchs2013_14_Ontario$agecatold <- cut(cchs2013_14_Ontario$DHH_AGE, c(12,30,40,50,60,70,80,105), 
                                  include.lowest = TRUE, right = FALSE)
levels(cchs2013_14_Ontario$agecatold) <- c("12-29","30-39","40-49","50-59","60-69","70-79","80+")

cchs2013_14_Ontario$agecat <- cut(cchs2013_14_Ontario$DHH_AGE, c(12,16,25,35,55,105), 
                                  include.lowest = TRUE, right = FALSE)
levels(cchs2013_14_Ontario$agecat) <- c("12-15","16-24","25-34","35-54","55+")

cchs2013_14_Ontario$educat <- factor(cchs2013_14_Ontario$EDUDR04, 
                                     labels=c("Less than secondary","Secondary school grad", 
                                              "Some post-secondary", "Post-secondary grad", "NS"),
                                     levels =c(1,2,3,4,9))

cchs2013_14_Ontario$incpercat <- with(cchs2013_14_Ontario,
                                      ifelse(INCDPER < 10 | INCDPER %in% c(96,99), INCDPER,
                                             ifelse(INCDPER %in% c(10,11), 10,
                                                    ifelse(INCDPER %in% c(12,13), 11, 
                                                           ifelse(INCDPER == 14, 12, NA)))))

cchs2013_14_Ontario$incpercat <- factor(cchs2013_14_Ontario$incpercat,
                                        labels = c("No income", "LESS THAN $5,000", "$5,000 TO $9,999", "$10,000 TO $14,999", "$15,000 TO $19,999", "$20,000 TO $29,999", "$30,000 TO $39,999", "$40,000 TO $49,999", "$50,000 TO $59,999", "$60,000 TO $79,999","$80,000 TO 99,999","$100k or more",'NA',"NS"),
                                        levels =c(1:12,96,99))

cchs2013_14_Ontario$inchhcat <- with(cchs2013_14_Ontario,
                                     ifelse(INCDHH < 10 | INCDHH %in% c(96,99), INCDHH,
                                            ifelse(INCDHH %in% c(10,11), 10,
                                                   ifelse(INCDHH %in% c(12,13), 11, 
                                                          ifelse(INCDHH %in% c(14,15), 12, NA)))))
cchs2013_14_Ontario$inchhcat <- factor(cchs2013_14_Ontario$inchhcat,
                                       labels = c("No income", "LESS THAN $5,000", "$5,000 TO $9,999", "$10,000 TO $14,999", "$15,000 TO $19,999", "$20,000 TO $29,999", "$30,000 TO $39,999", "$40,000 TO $49,999", "$50,000 TO $59,999", "$60,000 TO $79,999","$80,000 TO 99,999","$100k or more","NS"),
                                       levels =c(1:12,99))

cchs2013_14_Ontario$ethncat <- factor(cchs2013_14_Ontario$SDCDCGT,
                                      labels = c('White','Black','Korean','Filipino','Japanese','Chinese','South Asian','South East Asian','Arab','West Asian','Latin American','Other Racial or Cultural Origin','Multiple Racial/Cultural Origins','NA','NS'),
                                      levels = c(1:13,96,99))

cchs2013_14_Ontario$maritalcat <- factor(cchs2013_14_Ontario$DHH_MS,
                                         labels = c('Married','Common-Law','Widowed','Separated','Divorced','Single/Never Married','DK','Refusal','NS'),
                                         levels = c(1:6,97,98,99))

cchs2013_14_Ontario$geourcat <- factor(cchs2013_14_Ontario$GEODUR2,
                                       labels=c("Population Centre", "Rural"),
                                       levels=c(1,2))

cchs2013_14_Ontario$alcfreqcat <- factor(cchs2013_14_Ontario$ALC_2,
                                         labels = c('less than once a month','once a month','2 to 3 times a month','once a week','2 to 3 times a week','4 to 6 times a week','every day','NA','DK','Refusal','NS'),
                                         levels = c(1:7,96:99))

cchs2013_14_Ontario$alctypecat <- factor(cchs2013_14_Ontario$ALCDTTM,
                                         labels = c("Regular drinker","Occasional drinker","Did not drink in the last 12 months","NS"),
                                         levels = c(1,2,3,9))

cchs2013_14_Ontario$alchvycat <- with(cchs2013_14_Ontario,
                                      ifelse(ALC_3 %in% 1:2, 2, # 1 is heavy drinking
                                             ifelse(ALC_3 %in% 3:6, 1, # 2 did not have heaving drinking
                                                    ifelse(ALC_3 %in% 96:99, ALC_3, NA))))

cchs2013_14_Ontario$alchvycat <- factor(cchs2013_14_Ontario$alchvycat,
                                        labels = c("Heavy drinking in the past year at least once per month","Did not heavy drink in the past year","NA","DK","Refusal","NS"),
                                        levels = c(1:2,96:99))

```

# Implement survey design

```{r svy_design, echo=TRUE, message=FALSE, warning=FALSE}

# See link for instructions https://stackoverflow.com/questions/45642026/survey-weights-and-boostrap-wieghts-to-get-counts-and-cis

library(survey)
cchsDesign <- svrepdesign(data = cchs2013_14_Ontario,
                        weights = ~WTS_S,
                        repweights = "BSW[0-5]+",
                        type = "bootstrap",
                        combined.weights = TRUE,
                        mse = TRUE)

```


# Comparison of original and input data for microsimulation

CCHS baseline data with survey weights are compared with data with rounded weights used for microsimulations. Comparison is based on the proportion of people belonging to different categories of each variable of interest. The two are virtually the same, as observed in the table and barplots below. They are however usedful for determining whether some categories should be merged together due to very low proportions, such as household income categories of low income.

```{r baseline_comparison, fig.width=8, fig.height=7}

# Compare CCHS survey weighted proportion against simulation rounded proprotions

vars <- c('agecat', 'sex', 'educat', 'incpercat', 'inchhcat', 'ethncat', 'maritalcat', 'geourcat','alcfreqcat','alctypecat','alchvycat')
varnames <- c('Age','Sex','Education','Personal income','Household income','Ethnicity','Marital status','Geography','Alcohol drinking frequency','Type of alcohol drinker','Heavy drinking')

varPropList <- list()

for (v in vars) {
  cchsProp <- prop.table(
      svytable(~get(v),
               design = cchsDesign)) %>%
      data.frame
  names(cchsProp) <- c('cat','cchs_prop')

  simProp <- cchs2013_14_Ontario[,c(v,'WTS_S')] %>%
      mutate(wgt_rd = round(WTS_S)) %>%
      group_by(get(v)) %>%
      summarise(wgt_rd = sum(wgt_rd)) %>%
      mutate(tot_wgt = sum(wgt_rd),
             prop_wgt = wgt_rd/tot_wgt)
  names(simProp)[1] <- 'cat'
  
  cchsSimProp <- merge(cchsProp, simProp)
  cchsSimProp$var <- v
  cchsSimProp$vname <- varnames[match(v,vars)]
  
  varPropList[[v]] <- cchsSimProp %>%
    arrange(cat)
}

varProp <- do.call(rbind, varPropList)
varProp$var <- factor(varProp$var, levels=vars)

varProp <- varProp %>% 
  arrange(var)

```

```{r comparison_table}

# Print table
varProp[c('vname','cat','cchs_prop','prop_wgt')] %>%
  kable(row.names = FALSE, col.names = c('Variable','Category','CCHS weighted proportion','Rounded proportions for microsim.')) %>%
  kable_styling("striped")

```

```{r comparison_barplots}

# Barplots
for (v in vars) {
  par(mar = c(9, 3, 2, 0.5), mgp = c(1.8, 0.8, 0))
  to_plot <- varPropList[[v]][c('cchs_prop','prop_wgt')] %>% as.matrix %>% t
  
  vname <- varnames[match(v,vars)]
  
  if (v == 'sex') {
    barplot_func <- function() barplot(to_plot, beside = TRUE, 
                                       xaxt = 'n', yaxt = 'n', 
                                       main = vname , ylim = c(0, 0.6),
                                       col = c('lightskyblue','blue4'),
                                       border = c('lightskyblue','blue4'))
  } else {
    barplot_func <- function() barplot(to_plot, beside = TRUE, 
                                       xaxt = 'n', yaxt = 'n', 
                                       main = vname ,
                                       col = c('lightskyblue','blue4'),
                                       border = c('lightskyblue','blue4'))
  }
  x_at <- matrix(barplot_func(), ncol = 2, byrow = TRUE)
  x_at <- apply(x_at, 1, mean)
  
  x_labels <- gsub('(.{1,20})(\\s|$)', '\\1\n', as.character(varPropList[[v]]$cat))
  
  axis(side = 1, at = x_at, labels = x_labels, col='white', las=2)
  axis(side = 2, las = 2)
  
  legend_pos <- ifelse(v %in% c('ethncat','maritalcat','geourcat','alctypecat','alchvycat'),'topright','topleft')
  legend(legend_pos, legend = c('CCHS weighted','Rounded for microsim.'), 
         col = c('lightskyblue','blue4'), pch = 15, bty = 'n')
}

```

# Transition probabilities

```{r load_trans_prob}

transProb <- readxl::read_xlsx(path = file.path(dataDir, 'transition-prob-alc-micro-Hana.xlsx'),
                  sheet = 2, skip = 1)

transProb$sex <- gsub("M","Male",
                       gsub("F","Female", transProb$sex))

transProb$agecat <- factor(transProb$age)
transProb$percent_alcohol_attributable <- transProb$percent_alcohol_attributable/100

transProb[c('age','sex','transition_4_rate_100k','transition_3_rate_100k')] %>%
  kable(row.names = FALSE, col.names = c('Age','Sex','Alcoholic death rates (per 100K)','Non-alcoholic death rates (per 100K)')) %>%
  kable_styling("striped")

```

```{r data_list}

# Create a list with data.frame for the corresponding stratum in transProb
cchsDataList <- list()

for (i in 1:nrow(transProb)) {
  transProb_sub <- transProb[i,]
  select_agecat <- transProb_sub$agecat
  select_sex <- transProb_sub$sex
  
  if (select_agecat %in% c('Total','All')) {
    if (select_sex == 'All') {
      cchsDataList[[i]] <- cchs2013_14_Ontario # for all ages and both sex
    } else {
       cchsDataList[[i]] <- cchs2013_14_Ontario %>% 
         filter(sex == as.character(select_sex)) # for all ages and by sex
    }
  } else {
    cchsDataList[[i]] <- cchs2013_14_Ontario %>% 
      filter(agecat == as.character(select_agecat) & sex == as.character(select_sex)) # for specific age and sex
  }
}

rm(cchs2013_14_Ontario)

```


# Combine alcoholic and non-alcholic attributable transition probabilities

## Poor method

```{r flowchart_v1}

library(DiagrammeR)
grViz("digraph flowchart {
      # node definitions with substituted label text
      graph [layout = dot, rankdir = LR]
      node [fontname = Helvetica, shape = rectangle, style=filled]        
      tab1 [label = 'Alcoholic attr. mort. rate', color = orange]
      tab2 [label = 'Non-alcoholic attr. mort. rate', color = grey]
      tab3 [label = 'Sample either \n  alcoholic or non-alcoholic \n  rate based on \n % mort. alcohol-attr.', shape = diamond]
      tab4 [label = 'Monte Carlo \n simulation on the \n selected mort. rate', shape = diamond]
      tab5 [label = 'Alcoholic attr. mort.', color = pink]
      
      # edge definitions with the node IDs
      tab1 -> tab3;
      tab2 -> tab3;
      tab3 -> tab4;
      tab4 -> tab5
      }
      ")

```

This make not be correct, since it is possible that a person die from both alcoholic and non-alcoholic causes. Also not sure they alcoholic and non-alcoholic attributable rates are exclusive of each other.

# Better method

```{r flowchart_v2}

library(DiagrammeR)
grViz("digraph flowchart {
      # node definitions with substituted label text
      graph [layout = dot, rankdir = LR]
      node [fontname = Helvetica, shape = rectangle, style=filled]        
      tab1 [label = 'Alcoholic att. mort. rate', color = orange]
      tab2 [label = 'Non-alcoholic att. mort. rate', color = grey]
      tab3 [label = 'Monte Carlo \n simulation', shape = diamond, color = orange]
      tab4 [label = 'Monte Carlo \n simulation', shape = diamond, color = grey]
      tab5 [label = 'Combine Mone Carlo\n results (summation)', shape = diamond]
      tab6 [label = 'Alcoholic att. mort.', color = pink]
      
      # edge definitions with the node IDs
      tab1 -> tab3;
      tab2 -> tab4;
      tab3 -> tab5;
      tab4 -> tab5;
      tab5 -> tab6
      }

      ")

```

This maybe a better solution. However, if alcoholic and non-alcoholic attributable rates are NOT exclusive of each other, then the mortality maybe inflated.

<!-- Run 3 state model -->

```{r run_sim_3state, eval=FALSE}

# Load microsimulation functions
source('sim_functions_3state.R')

# Run microsimulation
comp.time <- list()

for (i in seq(cchsDataList)[-9:11]) {

  # Data subset
  cchsData <- cchsDataList[[i]]
  transProbSub <- transProb[i,]
  
  # Model input
  n.i   <- sum(round(cchsData$WTS_S))   # number of simulated individuals, since it is weighted sum it needs to be rounded to join back to CCHS data
  n.t   <- 30                    # time horizon, 30 cycles
  v.n   <- c("H","S","D")        # the model states: Healthy (H), Sick (S), Dead (D)
  n.s   <- length(v.n)           # the number of health states
  v.M_1 <- rep("H", n.i)         # everyone begins in the healthy state 
  d.e <- 0.03                    # equal discounting of costs and QALYs by 3%
  samplev.m <- 1                 # argument use for samplev() function 
  
  # Transition probabilities (per cycle)
  p.HD   <- transProbSub$transition_3_rate_100k/100000      # probability to die when healthy
  p.HS   <- transProbSub$transition_1_rate_100k/100000      # probability to become sick when healthy
  p.SH   <- transProbSub$transition_2_rate_100k/100000      # probability to become healthy when sick
  #rr.S   <- transProbMat[i,'rr.S'] %>% unlist             	# rate ratio of death when sick vs healthy
  #r.HD   <- -log(1 - p.HD) 	     # rate of death when healthy 
  #r.SD   <- rr.S * r.HD  	       # rate of death when sick
  #p.SD   <- 1 - exp(- r.SD)       # probability to die when sick
  p.SD  <- transProbSub$transition_4_rate_100k/100000
  rp.S  <- 0.2                     # increase of the mortality rate with every additional year being sick
  
  # QALY inputs
  u.H     <- 1              # utility when healthy 
  u.S     <- 0.75           # utility of sick individuals with every additional year being sick
  ru.S    <- 0.03           # decrease in utility of sick individuals with every additional year being sick/sicker
  v.x     <- runif(n.i, 0.95, 1.05) # vector capturing individuals' effect modifier at baseline
  
  # START SIMULATION
  cat('\n# Stratum', paste(transProbSub[c('age','sex')], collapse=':'),', n =',n.i, '\n')      # display the stratum
  
  p = Sys.time()
  sim <- MicroSim(v.M_1, n.i, n.t, v.n, samplev.m, X = v.x, d.e, seed = 100)  # run for treatment
  comp.time[[i]] = Sys.time() - p
  
  comp.time[[i]]
  
  # SAVE SIMULATION RESULTS TO FILE
  file_path <- file.path(dataDir,paste0('sim_',i,'.rds'))
  saveRDS(sim, file_path)
  
  rm(sim)
}


```

# Run two-state model

```{r run_sim_2state, echo=TRUE, eval=FALSE}

# Load microsimulation functions
source('sim_functions_2state.R')

# Set memory limit
memory.limit(size=30000)

# Run microsimulations
comp.time <- list()

for (i in seq(cchsDataList)[1:8]) {

  # Data subset
  cchsData <- cchsDataList[[i]]
  transProbSub <- transProb[i,]
  
  # Model input (convert this into a function, incorporate bootstrap weights as input and do some testing!)
  n.i   <- sum(round(cchsData$WTS_S))   # number of simulated individuals, since it is weighted sum it needs to be rounded to join back to CCHS data
  n.t   <- 30                    # time horizon, 30 cycles
  v.n   <- c("H","D")            # the model states: Healthy (H), Dead (D)
  n.s   <- length(v.n)           # the number of health states
  v.M_1 <- rep("H", n.i)         # everyone begins in the healthy state 
  d.e <- 0.03                    # equal discounting of costs and QALYs by 3%
  samplev.m <- 1                 # argument use for samplev() function 
  
  # Transition probabilities (per cycle)
  p.HD1   <- transProbSub$transition_4_rate_100k/100000      # probability to die when healthy (alcoholic attributable)
  p.HD2   <- transProbSub$transition_3_rate_100k/100000      # probability to die when healthy (non-alcoholic attributable)
  p.AA    <- transProbSub$percent_alcohol_attributable       # % mortalities alcohol-attributable
  
  # QALY inputs
  u.H     <- 1              # utility when healthy 
  v.x     <- runif(n.i, 0.95, 1.05) # vector capturing individuals' effect modifier at baseline
  
  # START SIMULATION
  cat('\n# Stratum', paste(transProbSub[c('age','sex')], collapse=':'),', n =',n.i, '\n')      # display the stratum
  
  p = Sys.time()
  sim <- MicroSimV2(v.M_1, n.i, n.t, v.n, samplev.m, X = v.x, d.e, seed = 100)  # run for treatment
  comp.time[[i]] = Sys.time() - p
  
  print(comp.time[[i]])
  
  # SAVE SIMULATION RESULTS TO FILE
  file_path <- file.path(dataDir,paste0('sim_',i,'_2state_v2.rds'))
  saveRDS(sim, file_path)
  
  rm(sim)
}


```

# Results from two-state model for males ages 16-24

```{r results_2state, cache=TRUE}

## Load results
i <- 3 # select stratum
age <- transProb[i,'age']
sex <- transProb[i,'sex']

wideToLong <- function(mat) {
  mat.dt <- as.data.table(mat)
  mat.dt$ind <- as.numeric(gsub("ind ","",row.names(mat.dt)))
  mat.dt.melt <- reshape2::melt(mat.dt, id.vars = "ind", variable.name = "cycle")
  mat.dt.melt$cycle <- as.numeric(gsub("cycle.","",mat.dt.melt$cycle))  
  return(mat.dt.melt)
}

longtablespath <- file.path(dataDir,paste0('sim_',i,'_2state_v2_longtables.rds'))
m.M12dtype.dt.melt.file <- file.path(dataDir, paste0('m.M12dtype.dt.melt_',i,'.rds'))

if (!file.exists(longtablespath)) {
  
  filefullpath <- file.path(dataDir,paste0('sim_',i,'_2state_v2.rds'))
  sim <- readRDS(filefullpath)
  
  # Output matrices
  library(data.table)

  # m.M <- sim$m.M
  # m.M1 <- sim$m.M1
  # m.M2 <- sim$m.M2
  # m.E <- sim$m.E
  # m.un1 <- sim$m.un1
  # m.un2 <- sim$m.un2
  # arr.p1 <- sim$arr.p1
  # arr.p2 <- sim$arr.p2
  # m.dtype <- sim$m.dtype
  #   
  n.i <- nrow(sim$m.M)
  
  m.M.dt.melt <- wideToLong(sim$m.M)
  m.M.dt.melt$state <- factor(m.M.dt.melt$value, levels=c("H","S","D"))
  
  m.M1.dt.melt <- wideToLong(sim$m.M1)
  m.M1.dt.melt$state <- factor(m.M1.dt.melt$value, levels=c("H","S","D"))
  
  m.M2.dt.melt <- wideToLong(sim$m.M2)
  m.M2.dt.melt$state <- factor(m.M2.dt.melt$value, levels=c("H","S","D"))
  
  m.dtype.dt.melt <- wideToLong(sim$m.dtype)

  m.un1.dt.melt <- wideToLong(sim$m.un1)
  
  m.un2.dt.melt <- wideToLong(sim$m.un2)

  m.E.dt.melt <- wideToLong(sim$m.E)

  longtables <- list(m.M.dt.melt = m.M.dt.melt, m.M1.dt.melt = m.M1.dt.melt, m.M2.dt.melt = m.M2.dt.melt, m.dtype.dt.melt = m.dtype.dt.melt,
                     m.un1.dt.melt = m.un1.dt.melt, m.un2.dt.melt = m.un2.dt.melt, m.E.dt.melt = m.E.dt.melt)
  
  saveRDS(longtables, longtablespath)
  
  rm(longtables, m.un1.dt.melt, m.un2.dt.melt, m.E.dt.melt)
    
  # Merge m.M, m.M1, and m.M2
  m.M12.dt.melt <- merge.data.table(m.M1.dt.melt, m.M2.dt.melt, by = c("ind","cycle"))
  names(m.M12.dt.melt) <- gsub('\\.y','.nonalcoholic',
                               gsub('\\.x','.alcoholic', names(m.M12.dt.melt)))
  
  m.M12dtype.dt.melt <- merge.data.table(
    merge.data.table(m.M.dt.melt, m.M12.dt.melt, by = c("ind","cycle")),
    m.dtype.dt.melt, by = c("ind","cycle"))
  
  names(m.M12dtype.dt.melt) <- gsub('\\.y','.attributable',
                               gsub('\\.x','.combined', names(m.M12dtype.dt.melt)))
  
  saveRDS(m.M12dtype.dt.melt, m.M12dtype.dt.melt.file)
  
  rm(m.M.dt.melt, m.M1.dt.melt, m.M2.dt.melt, m.dtype.dt.melt, m.M12dtype.dt.melt)
  
  gc()
  
} else {
  longtables  <- readRDS(longtablespath)
  m.M12dtype.dt.melt <- readRDS(m.M12dtype.dt.melt.file)
  
  m.M.dt.melt = longtables$m.M.dt.melt
  m.M1.dt.melt = longtables$m.M1.dt.melt
  m.M2.dt.melt = longtables$m.M2.dt.melt
  m.dtype.dt.melt = longtables$m.dtype.dt.melt
  m.un1.dt.melt = longtables$m.un1.dt.melt
  m.un2.dt.melt = longtables$m.un2.dt.melt
  m.E.dt.melt = longtables$m.E.dt.melt
  
  gc()
}

longtablessubpath <- file.path(dataDir,paste0('sim_',i,'_2state_v2_longtables_sub.rds'))

if (!file.exists(longtablessubpath)) {
  
  # Subset to selected individuals
  ind.d <- m.M.dt.melt %>% filter(cycle == 30 & state == "D") %>% select(ind) %>% unlist
  
  set.seed(100)
  #n.i <- length(unique(m.M.dt.melt$ind))
  #select.ind <- sample(1:n.i, size = 100)
  select.ind <- sample(ind.d, size = 100)
  
  m.M.dt.melt.sub <- m.M.dt.melt %>% filter(ind %in% select.ind )
  m.M.dt.melt.sub$idx <- match(m.M.dt.melt.sub$ind, select.ind)
  
  m.M1.dt.melt.sub <- m.M1.dt.melt %>% filter(ind %in% select.ind ) 
  m.M1.dt.melt.sub$idx <- match(m.M1.dt.melt.sub$ind, select.ind)
  
  m.M2.dt.melt.sub <- m.M2.dt.melt %>% filter(ind %in% select.ind)
  m.M2.dt.melt.sub$idx <- match(m.M2.dt.melt.sub$ind, select.ind)
  
  m.dtype.dt.melt.sub <- m.dtype.dt.melt %>% filter(ind %in% select.ind)
  m.dtype.dt.melt.sub$idx <- match(m.dtype.dt.melt.sub$ind, select.ind)
  
  m.un1.dt.melt.sub <- m.un1.dt.melt %>% filter(ind %in% select.ind)
  m.un1.dt.melt.sub$idx <- match(m.un1.dt.melt.sub$ind, select.ind)
  
  m.un2.dt.melt.sub <- m.un2.dt.melt %>% filter(ind %in% select.ind)
  m.un2.dt.melt.sub$idx <- match(m.un2.dt.melt.sub$ind, select.ind)
  
  m.E.dt.melt.sub <- m.E.dt.melt %>% filter(ind %in% select.ind)
  m.E.dt.melt.sub$idx <- match(m.E.dt.melt.sub$ind, select.ind)
  
  longtablessub <- list(m.M.dt.melt.sub = m.M.dt.melt.sub, m.M1.dt.melt.sub = m.M1.dt.melt.sub, 
                        m.M2.dt.melt = m.M2.dt.melt.sub, m.dtype.dt.melt.sub = m.dtype.dt.melt.sub,
                     m.un1.dt.melt.sub = m.un1.dt.melt.sub, m.un2.dt.melt.sub = m.un2.dt.melt.sub,
                     m.E.dt.melt.sub = m.E.dt.melt.sub)
  
  saveRDS(longtablessub, longtablessubpath)
  
  gc()
} else {
  longtablessub  <- readRDS(longtablessubpath)
  
  m.M.dt.melt.sub = longtablessub$m.M.dt.melt.sub
  m.M1.dt.melt.sub = longtablessub$m.M1.dt.melt.sub
  m.M2.dt.melt.sub = longtablessub$m.M2.dt.melt.sub
  m.dtype.dt.melt.sub = longtablessub$m.dtype.dt.melt.sub
  m.un1.dt.melt.sub = longtablessub$m.un1.dt.melt.sub
  m.un2.dt.melt.sub = longtablessub$m.un2.dt.melt.sub
  m.E.dt.melt.sub = longtablessub$m.E.dt.melt.sub
  
  gc()
}

```

Heat maps showing health state of selected individuals (n=100), the randomly generated uniform numbers to determine transition to death or not, and the quality-adjusted life years (healthy = 1, death = 0).

```{r results_2state_heatmap, fig.width=6, fig.height=6}

# Plot heat maps
library(ggplot2)
ggplot(m.M.dt.melt.sub, aes(cycle, idx)) + 
  geom_tile(aes(fill = state), colour = "white")  +   
  scale_fill_manual(values= c("darkgreen","purple","red")) +
  labs(title = paste0('Age = ', age, ', sex = ',sex), 
       x = "cycle", y = "ind") +
  theme(text = element_text(size = 20)) 

ggplot(m.un1.dt.melt.sub, aes(cycle, idx)) +
  geom_tile(aes(fill = value)) +
    scale_fill_gradient2(low = "navy", mid = "white", high = "red",
                       midpoint = 0.5, na.value = "grey50", guide = "colourbar", aesthetics = "fill") +
  labs(fill = 'Randomly sampled uniform\nnumber for "alcoholic" portion\n(extremely high value would\ntransition to death)')

ggplot(m.un2.dt.melt.sub, aes(cycle, idx)) +
  geom_tile(aes(fill = value)) +
    scale_fill_gradient2(low = "navy", mid = "white", high = "red",
                       midpoint = 0.5, na.value = "grey50", guide = "colourbar", aesthetics = "fill") +
  labs(fill = 'Randomly sampled uniform\nnumber for "non-alcoholic" portion\n(extremely high value would\ntransition to death)')

ggplot(m.E.dt.melt.sub, aes(cycle, idx)) + 
  geom_tile(aes(fill = value)) + 
  labs(fill = 'Quality-adjusted life years')

```

Table showing the attributable cause of death among the selected individuals who died:

```{r results_2state_attributable, cache=TRUE}

# Merge m.M, m.M1, and m.M2
m.M12.dt.melt <- merge.data.table(m.M1.dt.melt, m.M2.dt.melt, by = c("ind","cycle"))
names(m.M12.dt.melt) <- gsub('\\.y','.nonalcoholic',
                             gsub('\\.x','.alcoholic', names(m.M12.dt.melt)))

m.M12dtype.dt.melt <- merge.data.table(
  merge.data.table(m.M.dt.melt, m.M12.dt.melt, by = c("ind","cycle")),
  m.dtype.dt.melt, by = c("ind","cycle"))

names(m.M12dtype.dt.melt) <- gsub('\\.y','.attributable',
                             gsub('\\.x','.combined', names(m.M12dtype.dt.melt)))
  
if (FALSE) {
  # check number of death by cause
  m.M12dtype.dt.melt[c("ind","cycle","value.alcoholic","value.nonalcoholic","value.attributable","value.combined")] %>%
  filter((value.alcoholic == "H" & value.nonalcoholic == "D") | (value.alcoholic == "D" & value.nonalcoholic == "H") | (value.alcoholic == "D" & value.nonalcoholic == "D")) %>%
  arrange(ind, cycle) %>%
  group_by(ind) %>%
  mutate(index = row_number()) %>%
  filter(index == 1) %>%
  mutate(cause = ifelse(value.alcoholic == "H" & value.nonalcoholic == "D", "Non-alcoholic",
                        ifelse(value.alcoholic == "D" & value.nonalcoholic == "H", "Alcoholic",
                               ifelse(value.alcoholic == "D" & value.nonalcoholic == "D", "Both", NA)))) %>%
  group_by(cause) %>%
  summarise(count = n())
  
  
  m.M12dtype.dt.melt %>% 
    filter((value.alcoholic == "H" & nonalcoholic == "D") | (value.alcoholic == "D" & value.nonalcoholic == "H")) # %>% # 13978
    #filter(cycle == 30) # 484
  
  m.M12dtype.dt.melt %>% 
    filter(value.alcoholic == "D" & value.nonalcoholic == "D") %>%
    group_by(ind) %>%
    mutate(row_index = row_number()) %>%
    filter(row_index == 1)
  
  sim$arr.p1[56,,6]
  sim$arr.p2[56,,6]
  
  sim$m.un1[56,6]
  sim$m.un2[56,6]
  
  sim$m.M1[56,]
  sim$m.M2[56,]
}

# Table for attributable cause of death
m.M12dtype.dt.melt[c("ind","cycle","value.alcoholic","value.nonalcoholic","value.attributable","value.combined")] %>%
  filter(ind %in% select.ind) %>%
  filter((value.alcoholic == "H" & value.nonalcoholic == "D") | (value.alcoholic == "D" & value.nonalcoholic == "H")) %>%
  kable(caption = "Attributable cause of death among the selected individuals who died",
        col.names = c("ind","cycle","Alcoholic path","Non-alcoholic path","Cause of death","Combined status"), row.names = FALSE) %>%
  kable_styling("striped")

if (FALSE) {
  select.death <- c(684132,774366)
}
  

```

Summary statistics of number of alcoholic/non-alcoholic deaths overall and by cycle for males ages 15-24:

```{r results_2state_summary, cache=TRUE}

# By individual
num_alc <- sum(apply(m.dtype == "Alcoholic", 1, any, na.rm=TRUE))
num_nonalc <- sum(apply(m.dtype == "Non-alcoholic", 1, any, na.rm=TRUE))
num_combined <- sum(apply(m.dtype, 1, 
          function(x) {
            any(x %in% c("Alcoholic","Non-alcoholic"), na.rm=TRUE)}))
percent_alc <- paste0(round(num_alc*100/n.i,1), '%')
percent_nonalc <- paste0(round(num_nonalc*100/n.i,1), '%')
percent_combined <- paste0(round(num_combined*100/n.i,1), '%')

summary <- data.frame(n.i = n.i, num_alc, num_nonalc, num_combined, 
                      percent_alc, percent_nonalc, percent_combined)

kable(summary, caption = "Number of alcoholic/non-alcoholic deaths overall", 
      col.names=c("Number of individual","Number of alcoholic deaths","Number of non-alcoholic deaths",
                           "Number of deaths combined", "Percent of alcoholic deaths", "Percent of non-alcoholic deaths", 
                           "Percent of deaths combined"),
      row.names = FALSE) %>%
  kable_styling("striped")

# By cycle
num_alc_by_cycle <- apply(m.dtype == "Alcoholic", 2, sum, na.rm=TRUE)
num_nonalc_by_cycle <- apply(m.dtype == "Non-alcoholic", 2, sum, na.rm=TRUE)
num_combined_by_cycle <- apply(m.dtype, 2, 
          function(x) {
            sum(x %in% c("Alcoholic","Non-alcoholic"), na.rm=TRUE)})

percent_alc_by_cycle <- paste0(round(num_alc_by_cycle*100/n.i,1), '%')
percent_nonalc_by_cycle <- paste0(round(num_nonalc_by_cycle*100/n.i,1), '%')
percent_combined_by_cycle <- paste0(round(num_combined_by_cycle*100/n.i,1), '%')

summary_by_cycle <- data.frame(num_alc_by_cycle, num_nonalc_by_cycle, num_combined_by_cycle, 
                      percent_alc_by_cycle, percent_nonalc_by_cycle, percent_combined_by_cycle)

kable(summary_by_cycle, caption = "Number of alcoholic/non-alcoholic deaths by cycle", 
      col.names=c("Number of alcoholic deaths","Number of non-alcoholic deaths",
                           "Number of deaths combined", "Percent of alcoholic deaths", "Percent of non-alcoholic deaths", 
                           "Percent of deaths combined"),
      row.names = FALSE) %>%
  kable_styling("striped")

```

```{r ind_results_2state_v1, eval=FALSE}

## Individual level results
# Select data from the first stratum
cchsData <- cchsDataList[[i]]

# Survey weights for expanding CCHS data
cchs_wgts <- round(cchsData$WTS_S)
cchs_nrow <- length(cchs_wgts)
expand_wgts_list <- sapply(1:cchs_nrow, function (i) {
  rep(i, times=cchs_wgts[i])
})
expand_wgts <- unlist(expand_wgts_list)

# Expand data according to survey weights
cchs_sub_weighted <- cchsData[expand_wgts,
                         c('agecat','sex','educat','incpercat','inchhcat',
                           'ethncat','maritalcat','geourcat')]

# Calculate number of healthy and sick health states
num_hlth_sick_states <- apply(m.M, 1, 
      function(i) {
        table(i)[c("H","S")]
      }) 
num_hlth_sick_states <- data.frame(num_hlth_sick_states)
cchs_sub_weighted[c("num_hlth_state","num_sick_state")] <- t(num_hlth_sick_states)

# Time of death and Assign health state at the end of cycles
death_time_cycles_end_list <- apply(m.M, 1, 
                               function(i) {
                                 
                                 d_times <- grep("D",i)
                                 
                                 if (length(d_times) != 0) {
                                   
                                   min_pos <- min(grep("D",i))
                                   time_of_death <- min_pos 
                                 } else {
                                   time_of_death <- NA 
                                 }
                                 
                                 end_of_cycles <- tail(i, 1)
                                 
                                 return(
                                   list(time_of_death = time_of_death, end_of_cycles=end_of_cycles))
                               }) 
death_time_cycles_end <- do.call(rbind, death_time_cycles_end_list)
death_time_cycles_end <- data.frame(death_time_cycles_end)
cchs_sub_weighted[c('time_of_death','end_of_cycles')] <- death_time_cycles_end

# QALYs per individual
cchs_sub_weighted$qalys <- rowSums(m.E) 

# Show the results
DT::datatable(cchs_sub_weighted[select.ind,], 
              colnames=c("Age group","Sex","Education","Income (personal)","Income (household)",
                         "Ethnicity","Marital Status","Urban/Rural", 
                         "Number of healthy state","Number of sick state",
                         'Time of death','End of cycles',
                         "QALYs"))

```


# Percent of deaths over time, for all age- and sex-groups

```{r results_overtime, cache=TRUE}

## Load results
m.dtype.cycle.wide.list <- list()
m.dtot.cycle.list <- list()
n.i.list <- list()

dtype <- c("Alcoholic","Non-alcoholic","Both")
dtype.df <- data.frame(x = dtype)

for (i in 1:8) {
  
  filefullpath <- file.path(dataDir,paste0('sim_',i,'_2state_v2.rds'))
  sim <- readRDS(filefullpath)
  
  m.dtype <- sim$m.dtype
  n.i <- nrow(m.dtype)
  
  m.dtype.cycle.list <- apply(m.dtype, 2, function(x) {
    
    dtype.tab <- table(x)
    
    if (length(dtype.tab) != 0){
      df <- data.frame(dtype.tab)
      dtype.df.mg <- merge(dtype.df, df, by='x', all.x=TRUE)
      dtype.df.mg$Freq <- with(dtype.df.mg, ifelse(is.na(Freq), 0, Freq))
    } else {
      dtype.df.mg <- dtype.df
      dtype.df.mg$Freq <- 0
    }
    
    return(dtype.df.mg)
  })
  
  for (x in seq(m.dtype.cycle.list)) {
    m.dtype.cycle.list[[x]]$cycle <- x
  }
  
  m.dtype.cycle <- do.call(rbind, m.dtype.cycle.list)
  m.dtype.cycle.wide <- reshape(m.dtype.cycle, idvar = "cycle", timevar = "x", 
                                direction = "wide")
  
  m.dtype.cycle.wide$Percent.Alcoholic <- m.dtype.cycle.wide$Freq.Alcoholic*100 / n.i
  m.dtype.cycle.wide$"Percent.Non-alcoholic" <- m.dtype.cycle.wide$"Freq.Non-alcoholic"*100 / n.i
  m.dtype.cycle.wide$Percent.Both <- m.dtype.cycle.wide$Freq.Both*100 / n.i
  
  m.dtot.cycle <- m.dtype.cycle %>% 
    group_by(cycle) %>%
    summarise(Freq = sum(Freq)) %>%
    mutate(Percent = Freq*100/n.i)
  
  plot(m.dtot.cycle$cycle, m.dtot.cycle$Percent, type='l', yaxt='n',
       xlab = 'Cycle', ylab = 'Percent of deaths (%)')
  axis(2, las = 2)
  
  n.i.list[[i]] <- n.i
  m.dtype.cycle.wide.list[[i]] <- m.dtype.cycle.wide
  m.dtot.cycle.list[[i]] <- m.dtot.cycle
  
  rm(sim, m.dtype)
}

```

```{r results_overtime_allcause}

## Plots
# All-cause
transProb$lty <- as.numeric(as.factor(transProb$sex))
transProb$col <- with(transProb, 
                      ifelse(age == '16-24', 'red',
                             ifelse(age == '25-34', 'orange',
                                    ifelse(age == '35-54', 'green',
                                           ifelse(age == '55+', 'blue', NA)))))


par(mar=c(4,4,1,4), mgp=c(1.8,0.8,0), cex=0.9)

plot(0, 0, type = 'n', xlim=c(1,31), ylim = c(0,7), yaxt='n',
     xlab = 'Cycle', ylab = "Percent of deaths ages 16-54 (%)", )
axis(2, las=2)

for (i in 1:6) {
  lty <- transProb[i,'lty'] %>% unlist
  col <- transProb[i,'col'] %>% unlist
  
  lines(m.dtot.cycle.list[[i]]$cycle, m.dtot.cycle.list[[i]]$Percent, 
       lty = lty, lwd = 2, col = col)
}

par(new = TRUE)
i = 7
lty <- transProb[i,'lty'] %>% unlist
col <- transProb[i,'col'] %>% unlist
plot(m.dtot.cycle.list[[i]]$cycle, m.dtot.cycle.list[[i]]$Percent, 
     type = "l", axes = FALSE, yaxt='n', bty = "n", xlab = "", ylab = "",
     ylim = c(0,60), lty = lty, lwd = 2, col = col)
axis(side = 4, at = pretty(range(m.dtot.cycle.list[[i]]$Percent)), las=2)
mtext("Percent of deaths ages 55+ (%)", side=4, line=2, cex=0.9)
i = 8
lty <- transProb[i,'lty'] %>% unlist
col <- transProb[i,'col'] %>% unlist
lines(m.dtot.cycle.list[[i]]$cycle, m.dtot.cycle.list[[i]]$Percent,
      lty = lty, lwd = 2, col = col)

legend('topleft', legend = c('Ages 16-24 (left-axis)','Ages 25-34 (left-axis)', 'Ages 35-54 (left-axis)',
                             'Ages 55+ (right-axis)','Male','Female'),
       col = c('red','orange','green','blue',grey(0.4),grey(0.4)),
       lty = c(1,1,1,1,3,2), lwd = 2)


```

```{r results_overtime_cause_specific_plot1, fig.width = 12, fig.height= 16}

# Cause-specific
par(mfrow=c(4,2), mar=c(4,5,1.5,4), mgp=c(2.5,0.8,0), cex=1.0)

for (i in 1:8) {
  age <- transProb[i,'age'] %>% unlist
  sex <- transProb[i,'sex'] %>% unlist
  
  m.dtype.cycle.wide <- m.dtype.cycle.wide.list[[i]]
  plot(m.dtype.cycle.wide$cycle, m.dtype.cycle.wide$Percent.Alcoholic,
       yaxt='n', ylim=c(0,1.4), pch=16, col='red',
       main = paste0(sex,', ages ',age,' (n = ',n.i.list[[i]],')'),
         xlab='Cycle', ylab="Percent of alcoholic deaths (%)") 
  axis(2, las=2)
  
  if (i %in% 1:6) {
    ylim = c(0,6)
  } else {
    ylim = c(0,60)
  }
  
  par(new = TRUE)
  plot(m.dtype.cycle.wide$cycle, m.dtype.cycle.wide$"Percent.Non-alcoholic", 
     type = "l", lwd = 2, axes = FALSE, yaxt='n', bty = "n", ylim = c(ylim[1],ylim[2]),
     xlab = "", ylab = "",
     lty = 2, col = 'black')
  if (i %in% 1:6) {
    axis(side = 4, at = 0:5, las=2)
  } else {
    axis(side = 4, at = seq(0,60,by=10), las=2)
  }
  mtext("Percent of non-alcoholic deaths (%)", side=4, line=1.8, cex=1.0)
  
  legend('topleft', legend=c('Alcoholic (left-axis)','Non-alcoholic (right-axis)'), pch = c(16,NA), lty = c(NA, 2), col=c('red','black'))
}

 
```


```{r results_overtime_cause_specific_plot2, fig.width = 12, fig.height= 8}

## Plots
# Alcohol
transProb$lty <- as.numeric(as.factor(transProb$sex))
transProb$col <- with(transProb, 
                      ifelse(age == '16-24', 'red',
                             ifelse(age == '25-34', 'orange',
                                    ifelse(age == '35-54', 'green',
                                           ifelse(age == '55+', 'blue', NA)))))


par(mfrow = c(2,2), mar=c(4,4,1,0.5), mgp=c(2.5,0.8,0), cex=0.9)

# Alcoholic deaths ages 16-54
plot(0, 0, type = 'n', bty='n', 
     xlim=c(1,31), ylim = c(0,1.5), yaxt='n',
     main = 'Alcoholic deaths ages 16-54',
     xlab = 'Cycle', ylab = "Percent of alcoholic deaths (%)", )
axis(2, las=2)

for (i in 1:6) {
  lty <- transProb[i,'lty'] %>% unlist
  col <- transProb[i,'col'] %>% unlist
  
  m.dtype.cycle.wide <- m.dtype.cycle.wide.list[[i]]
  
  lines(m.dtype.cycle.wide$cycle, m.dtype.cycle.wide$Percent.Alcoholic,
  lwd = 2, lty = lty, col=col) 
}
  
legend('topleft', legend = c('Ages 16-24','Ages 25-34', 'Ages 35-54',
                             'Ages 55+','Male','Female'),
       col = c('red','orange','green','blue',grey(0.4),grey(0.4)),
       lwd = 2, lty = c(1,1,1,1,3,2), inset = c(0,0.05))

# Alcoholic deaths ages 55+
plot(0, 0, type = 'n', bty='n', 
     xlim=c(1,31), ylim = c(0,1.5), yaxt='n',
     main = 'Alcoholic deaths ages 55+',
     xlab = 'Cycle', ylab = "Percent of alcoholic deaths (%)")
axis(2, las=2)

for (i in 7:8) {
  lty <- transProb[i,'lty'] %>% unlist
  col <- transProb[i,'col'] %>% unlist
  
  m.dtype.cycle.wide <- m.dtype.cycle.wide.list[[i]]
  
  lines(m.dtype.cycle.wide$cycle, m.dtype.cycle.wide$Percent.Alcoholic,
  lwd = 2, lty = lty, col=col) 
}
  

# Non-alcoholic deaths ages 16-54
plot(0, 0, type = 'n', bty='n', 
     xlim=c(1,31), ylim = c(0,5), yaxt='n',
     main = 'Non-alcoholic deaths ages 16-54',
     xlab = 'Cycle', ylab = "Percent of non-alcoholic deaths (%)")
axis(2, las=2)

for (i in 1:6) {
  lty <- transProb[i,'lty'] %>% unlist
  col <- transProb[i,'col'] %>% unlist
  
  m.dtype.cycle.wide <- m.dtype.cycle.wide.list[[i]]
  
  lines(m.dtype.cycle.wide$cycle, m.dtype.cycle.wide$'Percent.Non-alcoholic',
  lwd = 2, lty = lty, col=col) 
}

#  Non-alcoholic deaths ages 55+
plot(0, 0, type = 'n', bty='n', 
     xlim=c(1,31), ylim = c(0,60), yaxt='n',
     main = 'Non-alcoholic deaths ages 55+',
     xlab = 'Cycle', ylab = "Percent of non-alcoholic deaths (%)")
axis(2, las=2)

for (i in 7:8) {
  lty <- transProb[i,'lty'] %>% unlist
  col <- transProb[i,'col'] %>% unlist
  
  m.dtype.cycle.wide <- m.dtype.cycle.wide.list[[i]]
  
  lines(m.dtype.cycle.wide$cycle, m.dtype.cycle.wide$'Percent.Non-alcoholic',
  lwd = 2, lty = lty, col=col) 
}
  

```


